Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-25 appendix.qmd
working on fig 7

@@

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation."
#| out-width: 30%
#| fig-pos: H

@@

Points are assigned to the bin they fall into based on the nearest centroid. If a point is equidistant from multiple centroids, it is assigned to the centroid with the lowest hexagonal bin ID.

@@

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the lowest centroid."
#| fig-width: 9
#| fig-height: 6
#| fig-pos: H

@@

Following the process in @xia2023 again using the Human Peripheral Blood Mononuclear Cells (PBMC) data. Note that @xia2023 uses a different subset from the PBMC dataset used by @chen2024, which contains $31,021$ cells including cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

>>>>

## Generating the nonlinear clusters data

@@

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation."
#| out-width: 30%
#| fig-pos: H
#| eval: false

@@

Points are assigned to the bin they fall into based on the nearest centroid. If a point is equidistant from multiple centroids, it is assigned to the centroid with the smallest bin ID.

@@

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the centroid with the smallest bin ID."
#| fig-width: 9
#| fig-height: 6
#| fig-pos: H

@@

Note that @xia2023 uses a different subset from the PBMC dataset used by @chen2024, shown by us in the main paper example, which is why this comparison is shown here and not in the main paper. Their data contains $31,021$ cells including cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell. Pre-processing follows the process in @xia2023 again using the Human Peripheral Blood Mononuclear Cells (PBMC) data.

### 2025-06-25 paper.qmd
working on fig 7

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. Both tNSE and UMAP can be considered to produce the \kD{} representation by minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

@wickham2015 provides several examples of models overlaid on the data in \pD{}. In hierarchical clustering, a representation of the dendrogrom using points and lines can be constructed by augmenting the data with points marking merging of clusters. Showing the movie of linear projections reveals shows how the algorithm sequentially fitted the cluster model to the data. For linear discriminant analysis or model-based clustering the model can be indicated by $(p-1)\text{-}D$ ellipses. It is possible to see whether the elliptical shapes appropriately matches the variance of the relevant clusters, and to compare and contrast different fits. For PCA, one can display the \kD{} plane of the reduced dimension using wireframes of transformed cubes. Using a wireframe is the approach we take here, to represent the NLDR model in \pD{}.

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@joreskog1969] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with hyper-parameters $\mathbfit{\theta}$. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

error_plot_two_curvy_quasi <- ggplot(
  error_df_two_curvy_abs, 
  aes(x = sqrt_row_wise_total_error, 
      y = 0, 
      colour = error_cat)) +
  geom_quasirandom(alpha = 0.5) +
  interior_annotation(label = "a", cex = 1.5) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab(expression(e[hj])) +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 0, 0))
```

@@

#### Computing hexagon grid configuration

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighboring observations, in both the NLDR layout (@fig-NLDR-two-curvy (d)) and the \pD{} model representation. When the data has been binned the triangulation connects centroids. The edges preserve the neighborhood information from the \kD{} representation when the model is lifted into \pD{}.

@@

The last step is to lift the \gD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. We use the \pD{} mean of the points in a given hexagon, $H_h$, denoted $C_{h}^{(p)}$, to map the centroid $C_{h}^{(2)} = (c_{h1}, c_{h2})$ to a point in \pD{}. Let the \pD{} mean be

$$C_{h}^{(p)} = \frac{1}{n_h}\sum_{i =1}^{n_h} x_i, h = {1, \dots, b; n_h > 0}.$$

@@

The model here is similar to a confirmatory factor analysis model [@brown2015], $\widehat{T}(X_1, X_2, X_3, X_4) + \Epsilon$. The difference between the fitted model and observed values would be considered to be residuals, and for this problem are \fD{}.

@@

$$\sqrt{\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$${#eq-equation1}

where $n$ is the number of observations, $b$ is the number of bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables and $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon. We can consider $e_{hj} = \sqrt{\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$ to be the residual for each observation.

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "Examining the distribution of residuals in a jittered dotplot (a), \\gD{} NLDR layout (b) and a tour of \\fD{} data space (c). Color indicates error ($e_{hj}$), dark color indicating high error and light indicates low error. Most large errors are distributed in one cluster (bottom one in c) and most small errors are distributed in the other cluster."
#| out-width: 100%
#| fig-pos: H
#| fig-width: 8
#| fig-height: 3

@@

A new benefit of this fitted model is that it allows us to now predict a new observation's value in the NLDR, for any method. The steps are to determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$ and predict it to be the centroid of this bin in \gD{}, $C^{(2)}_{h}$. This can be written as, let $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$, then the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$.

@@

The model fitting can be adjusted using these parameters:

- hexagon bin parameters
    - bottom left bin position $(s_1, \ s_2)$, 
    - the total number of bins ($b$), 
- bin density cutoff, to remove low-density hexagons.

@@

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-pos: H
#| fig-cap: "Hexbin density plots of tSNE layout of the nonlinear cluster data, using three different bin inputs: (a) $b = 240/98 \\text{ } (15, \\text{ }16)$, (b) $b = 720/215 \\text{ } (24, \\text{ }30)$, and (c) $b = 2496/549 \\text{ } (48, \\text{ }52)$. Color indicates standardized counts, dark indicating high count and light indicates low count. At the smallest bin size, the data structure is discontinuous, suggesting that there are too many bins. Using the RMSE of the model fit in $7\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 9
#| fig-height: 3

@@

#### Measurement of capturing the data shape in \gD{}

<!-- I'll be honest I've no idea what this section is -->
The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.

#### Removal of low density bins

By default, when assessing the choice of $b_1$, the total number of bins is measured by the number of **non-empty** bins. This more accurately reflects the hexagon grid relative the RMSE than the full number of bins in the grid. It may also be beneficial to remove low count bins also, in the situation where data is clustered or stringy, where the observed data is sparse. In order to decide if this is necessary, you would examine the distribution of bin counts, or the density which puts the counts on a standard scale. If there is something of a gap at low values, this would suggest a potential value to use as a cutoff. Alternatively, one could choose to remove based on a percentile, the bins with density in the lowest $5\%$ of all bins, for example. @fig-param-two-curvy (c) illustrates the effect on the model representation of removing bins below different percentages. Generally, we would urge caution in removing low count bins.

@@

```{r}
#| label: rm-lwd-bin-error

@@

```{r}
#| label: read-two-curvy-clust-nldr
# Read a variety of different NLDR representations of two_non_linear_diff_shaped_close_clusters
# and plot them on same aspect ratio
tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_47.rds")

>>>>

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. Both tNSE and UMAP can be considered to produce the \kD{} representation by minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE are examples of diffusion processes spreading to capture geometric shapes, that include both global and local structure. (See @coifman2005 for an explanation of diffusion processes.) TriMAP and PaCMAP

@@

@wickham2015 provides several examples of models overlaid on the data in \pD{}. In hierarchical clustering, a representation of the dendrogrom using points and lines can be constructed by augmenting the data with points marking merging of clusters. Showing the movie of linear projections reveals shows how the algorithm sequentially fitted the cluster model to the data. For linear discriminant analysis or model-based clustering the model can be indicated by $(p-1)\text{-}D$ ellipses. It is possible to see whether the elliptical shapes appropriately matches the variance of the relevant clusters, and to compare and contrast different fits. For PCA, one can display the model (a \kD{} plane of the reduced dimension) using wireframes of transformed cubes. Using a wireframe is the approach we take here, to represent the NLDR model in \pD{}.

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{} NLDR layout is the simplified representation. Typically, $k=2$, which is used for the rest of this paper. From a statistical perspective we can consider the distances between points in the \gD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@joreskog1969] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times 2}$, with hyper-parameters $\mathbfit{\theta}$. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

error_plot_two_curvy_quasi <- ggplot(
  error_df_two_curvy_abs, 
  aes(x = sqrt_row_wise_total_error, 
      y = 0, 
      colour = error_cat)) +
  geom_quasirandom(alpha = 0.5) +
  interior_annotation(label = "a", cex = 1.5) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab(expression(e[hi])) +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 0, 0))
```

@@

#### Hexagon grid configuration

@@

Delaunay triangulation [@lee1980;@alb2024] is used to connect points so that edges indicate neighboring observations, in both the NLDR layout (@fig-NLDR-two-curvy (d)) and the \pD{} model representation. When the data has been binned the triangulation connects centroids. The edges preserve the neighborhood information from the \gD{} representation when the model is lifted into \pD{}.

@@

The last step is to lift the \gD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. We use the \pD{} mean of the points in a given hexagon, $H_h$, denoted $C_{h}^{(p)}$, to map the centroid $C_{h}^{(2)} = (c_{h1}, c_{h2})$ to a point in \pD{}. Let the $j^{th}$ component of the \pD{} mean be

$$C_{hj}^{(p)} = \frac{1}{n_h}\sum_{i =1}^{n_h} x_{hij}, ~~~h = 1, \dots, b;~ j=1, \dots, p; ~ n_h > 0.$$

@@

The model here is similar to a confirmatory factor analysis model [@brown2015], $\widehat{T} + \Epsilon$. The difference between the fitted model and observed values would be considered to be residuals, and are \pD{}.

@@

$$\sqrt{\frac{1}{n}\sum_{h = 1}^{m}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$${#eq-equation1}

where $n$ is the number of observations, $m$ is the number of non-empty bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables and $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon. We can consider $e_{hi} = \sqrt{\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$ to be the residual for each observation.

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "Examining the distribution of residuals in a jittered dotplot (a), \\gD{} NLDR layout (b) and a tour of \\fD{} data space (c). Color indicates error ($e_{hi}$), dark color indicating high error and light indicates low error. Most large errors are distributed in one cluster (bottom one in c) and most small errors are distributed in the other cluster."
#| out-width: 100%
#| fig-pos: H
#| fig-width: 8
#| fig-height: 3

@@

A new benefit of this fitted model is that it allows us to now predict the NLDR value of a new observation, $x'$, for any method. The steps are to determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$ and predict it to be the centroid of this bin in \gD{}, $C^{(2)}_{h}$.

@@

The model fitting can be adjusted using these parameters: 
- hexagon bin parameters
    - bottom left bin position $(s_1, \ s_2)$, 
    - the number of bins in the horizontal direction ($b_1$), which controls the number of bins the vertical direction ($b_2$), total number of bins ($b$), and total number of non-empty bins ($m$).
- bin density cutoff, to possibly remove low-density hexagons.

@@

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-pos: H
#| fig-cap: "Hexbin density plots of tSNE layout of the nonlinear cluster data, using three different bin specifications ($b_1, b_2, b, m$): (a) 15, 16, 240, 98, (b) 24, 30, 720, 215, and (c) 48, 52, 2496, 549. Color indicates standardized counts, dark indicating high count and light indicates low count. At the smallest bin size, the data structure is discontinuous, suggesting that there are too many bins. Using the RMSE of the model fit in $7\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 9
#| fig-height: 3

@@

#### Handling of low density bins
Bin density is computed as $n_h/n$. To make comparisons between hexagon configurations made from different parameter choices we use the number of **non-empty** bins ($m$) rather than the total number of bins.

By default, when assessing the choice of horizontal bins, $b_1$, the total number of bins is measured by the number of **non-empty** bins ($m$). This more accurately reflects the hexagon grid relative to the RMSE than the full number of bins in the grid. It may also be beneficial to remove low count bins also, in the situation where data is clustered or stringy, where the observed data is sparse. In order to decide if this is necessary, you would examine the distribution of bin counts, or the density which puts the counts on a standard scale. If there is something of a gap at low values, this would suggest a potential value to use as a cutoff. Alternatively, one could choose to remove based on a percentile, the bins with density in the lowest $5\%$ of all bins, for example. @fig-param-two-curvy (c) illustrates the effect on the model representation of removing bins below different percentages.

@@

<!-- I'll be honest I've no idea what this section is -->
The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.
```{r}
#| label: rm-lwd-bin-error

@@

```{r}
#| label: read-two-curvy-clust-nldr
# Read a variety of different NLDR representations of two_non_linear_diff_shaped_close_clusters
# and plot them on same aspect ratio
tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds")
